### Task 2: Quality control
### created by Omar AlOmeir
### date: March 18, 2014

```{r}
library(RColorBrewer)
library(lattice)
library(preprocessCore)
library(impute)
```

Reading the data from the intake:
```{r}
dat <- read.table("data/GSE1710-data.tsv")
des <- readRDS("data/GSE1710-design.rds")
str(dat, max.level=0)
str(des)
```

heatmap of the first 100 rows before dealing with missing values:
```{r fig.width=7, fig.height=6}
mat <- as.matrix(dat)
bluesFun <- colorRampPalette(brewer.pal(n = 9, "Blues"))
heatmap(mat[1:100,], Rowv = NA, Colv = NA, col = bluesFun(256))
```

Dealing with missing values:
```{r}
# remove rows that contain more than 8 (25%) missing values
dat <- dat[rowSums(is.na(dat))<=8,]
str(dat, max.level=0)
```
We have `r nrow(dat)` rows left.

```{r}
# use knn to impute other NAs
mat <- as.matrix(dat)
mat <- impute.knn(mat)
dat <- as.data.frame(mat$data)
str(dat, max.level=0)
```

Heatmap of the first 100 rows:
```{r fig.width=7, fig.height=6}
mat <- as.matrix(dat)
bluesFun <- colorRampPalette(brewer.pal(n = 9, "Blues"))
heatmap(mat[1:100,], Rowv = NA, Colv = NA, col = bluesFun(256))
```

Smaples correlation heatmap:
```{r fig.width=7, fig.height=6}
sColors <- cm.colors(nrow(cor(mat)))
heatmap(cor(mat), Rowv = NA, Colv = NA, scale = "none", col = bluesFun(256), ColSideColors = sColors, RowSideColors = sColors, margins = c(10, 10))
```

Boxplot of means of correlation to check for outliers:
```{r fig.width=7, fig.height=6}
boxplot(apply(cor(mat), 2, mean))
which.min(apply(cor(mat), 2, mean))
```

GSM29606 has the lowest correlation of all, and can be removed as an outlier.
```{r fig.width=7, fig.height=6}
outlier <- subset(des, samples == "GSM29606")
olGroup <- subset(des, group == outlier$group & sex == outlier$sex)
splom(dat[olGroup$samples])
```

Removing the outlier:
```{r fig.width=7, fig.height=6}
roDes <- subset(des, samples != "GSM29606")
roDat <- dat[roDes$samples]
heatmap(cor(as.matrix(roDat)), Rowv = NA, Colv = NA, scale = "none", col = bluesFun(256))
boxplot(apply(cor(as.matrix(roDat)), 2, mean))
which.min(apply(cor(as.matrix(roDat)), 2, mean))
```

A boxplot of the data before normalization:
```{r fig.width=7, fig.height=6}
boxplot(roDat, pch = ".", ylab = "Expression level", las = 3)
```

Normalizing the data, using quantile normalization:
```{r}
nMat <- normalize.quantiles(as.matrix(roDat))
colnames(nMat) <- colnames(as.matrix(roDat))
rownames(nMat) <- rownames(as.matrix(roDat))
nDat <- as.data.frame(nMat)
```

A boxplot of the normalized data:
```{r fig.width=7, fig.height=6}
boxplot(nDat, pch = ".", ylab = "Expression level", las = 3)
```

Saving the normalized data without the outlier:
```{r}
write.table(nDat, "data/GSE1710-normalized-data.tsv")
saveRDS(roDes, "data/GSE1710-outlier-removed-design.rds")
```