### 01 Data intake
### Created by: Abrar Wafa
### Date: March 16, 2014


```{r}
library(limma)
library(lattice)
library(car)
library(RColorBrewer)
```

Read the data: 
```{r}
dat <- read.table("../../data/GSE1710-normalized-data.tsv")
des <- readRDS("../../data/GSE1710-outlier-removed-design.rds")
```


Separate age into groups:
```{r}
des$age <- as.numeric(levels(des$age))[as.integer(des$age)]
young <- which(des$age <= 30)
middleage <- which(des$age > 30 & des$age < 50)
old <- which(des$age >= 50)
des$age[young] <- "young"
des$age[middleage] <- "middleage"
des$age[old] <- "old"
des$age <- as.factor(des$age)
des$age <- recode(des$age, "",levels = c('young', 'middleage', 'old'))
str(des)
```


Create function to extract genes:
```{r}
prepareData <- function(g) {
  pDat0 <- data.frame()
  for (i in 1:length(g)) {
    pDat <- data.frame(des, gExp = as.vector(t(as.matrix(dat[g[i], ]))), 
                       gene = g[i])
    pDat0 <- rbind(pDat0, pDat)
  }
  pDat0
}
```

Set up design matrix to get 'reference + treatment effects' fit to test for group only:
i.e. testing: normal vs. CD, normal vs. UC, and UC vs. DC.
```{r}
desMat <- model.matrix(~group, des)
# Fitting two-way ANOVA to all probesets at once:
fit <- lmFit(dat, desMat)
ebFit <- eBayes(fit)
colnames(ebFit)
```

Use topTable to get the hits of all groups:
```{r}
hits <- topTable(ebFit, coef = grep("group", colnames(coef(ebFit))), number = nrow(dat), p.value=1e-2)
# save the resutls
write.table(hits, "data/groups-hits.tsv") 
str(hits) 
```
There are 1283 probes that have a BH-adjusted p-value less than 1e-2.


Explore hits of all groups:
```{r}
stripplot(gExp ~ group | gene, prepareData(head(rownames(hits), 6)), jitter.data = TRUE, auto.key = TRUE, type = c("p", "a"), grid = TRUE)
```

Explore non-hits of all groups:
```{r}
allprobes <- topTable(ebFit, coef = grep("group", colnames(coef(ebFit))), number = nrow(dat))
stripplot(gExp ~ group | gene, prepareData(tail(rownames(allprobes ), 6)), jitter.data = TRUE, auto.key = TRUE, type = c("p", "a"), grid = TRUE)
```

Display the top 50 probes in a heatmap:
```{r}
hit1 <- topTable(ebFit, coef = grep("group", colnames(coef(ebFit))), number = 50)
hGreys <- colorRampPalette(brewer.pal(n = 9, "Blues"))
hit1<- dat[rownames(hit1), ]
heatmap(as.matrix(hit1), Rowv = NA, Colv = NA, scale = "none", col = hGreys(256))
```

Test for any effect of CD group:
```{r}
CDhits <- topTable(ebFit, coef = grep("CD", colnames(coef(ebFit))), number = nrow(dat), p.value=1e-2)
write.table(CDhits, "data/CD-hits.tsv")
str(CDhits) 
```
There are 1453 probes that have a BH-adjusted p-value less than 1e-2.

Explore hits of CD groups:
```{r}
stripplot(gExp ~ group | gene, prepareData(head(rownames(CDhits), 6)), jitter.data = TRUE, auto.key = TRUE, type = c("p", "a"), grid = TRUE)
```

Display the top 50 probes in a heatmap:
```{r}
hits2 <- topTable(ebFit, coef = grep("CD", colnames(coef(ebFit))), number = 50)
hit2 <- dat[rownames(hits2), ]
heatmap(as.matrix(hit2), Rowv = NA, Colv = NA, scale = "none", col = hGreys(256))
```


Test for any effect of UC group:
```{r}
UChits <- topTable(ebFit, coef = grep("UC", colnames(coef(ebFit))), number = nrow(dat), p.value=1e-2)
str(UChits)
write.table(UChits, "data/UC-hits.tsv")
```
There are 527 probes that have a BH-adjusted p-value less than 1e-2.


Explore hits of UC groups:
```{r}
stripplot(gExp ~ group | gene, prepareData(head(rownames(UChits), 6)), jitter.data = TRUE, auto.key = TRUE, type = c("p", "a"), grid = TRUE)
```



Display the top 50 probes in a heatmap:
```{r}
hits3 <- topTable(ebFit, coef = grep("UC", colnames(coef(ebFit))), number = 50)
hit3 <- dat[rownames(hits3), ]
heatmap(as.matrix(hit3), Rowv = NA, Colv = NA, scale = "none", col = hGreys(256))
```


Using 'reference + treatment effects' fit to test for group and age:
```{r}
aMat <- model.matrix(~group*age, des)
afit <- lmFit(dat, aMat)
aebFit <- eBayes(afit)
colnames(aebFit)
```


Test for any effect of group and/or age:
```{r}
aghits <- topTable(aebFit, coef = c("groupCD", "groupUC", "agemiddleage", "ageold", "groupCD:agemiddleage", "groupUC:agemiddleage", "groupCD:ageold", "groupUC:ageold"), number = nrow(dat), p.value = 0.1)
str(aghits) 
```
There are 305 probes that have a BH-adjusted p-value less than 0.1.


Test for any effect of group only:
```{r}
ghits <- topTable(aebFit, coef = grep("group", colnames(coef(aebFit))),  number = nrow(dat), p.value = 0.1)
write.table(ghits, "data/groupVsage-hits.tsv")
str(ghits)
```
There are 921 probes that have a BH-adjusted p-value less than 0.1.


Test for any effect of age only:
```{r}
ahits <- topTable(aebFit, coef = grep("age", colnames(coef(aebFit))),  number = nrow(dat), p.value = 0.1)
str(ahits) 
```
There are only 4 probes that have a BH-adjusted p-value less than 0.1 which means there is no significant effect for age.

Explore hits of age only:
```{r}
stripplot(gExp ~ group | gene, prepareData(head(rownames(ahits), 4)), jitter.data = TRUE, auto.key = TRUE, type = c("p", "a"), grid = TRUE)
```

Use 'reference + treatment effects' fit to test for group and sex:
```{r}
sMat <- model.matrix(~group*sex, des)
sfit <- lmFit(dat, sMat)
sebFit <- eBayes(sfit)
colnames(sebFit)
```


Test for any effect of group and/or sex:
```{r}
sghits <- topTable(sebFit, coef = c("groupCD", "groupUC", "sexmale", "groupCD:sexmale", "groupUC:sexmale"), number = nrow(dat), p.value = 1e-2)
str(sghits)
```
There are 244 probes that have a BH-adjusted p-value less than 1e-2.

Test for any effect of group only:
```{r}
gHits <- topTable(sebFit, coef = grep("group", colnames(coef(sebFit))),  number = nrow(dat), p.value = 1e-2)
str(gHits)
```
There are 417 probes that have a BH-adjusted p-value less than 1e-2.


Test for any effect of sex:
```{r}
shits <- topTable(sebFit, coef = grep("sex", colnames(coef(sebFit))),  number = nrow(dat), p.value = 1e-2)
str(shits)
```
There is only 1 probe that have a BH-adjusted p-value less than 1e-2 which means there is no significant effect for sex.

Explore hits of sex only:
```{r}
stripplot(gExp ~ group | gene, prepareData(head(rownames(ahits), 1)), jitter.data = TRUE, auto.key = TRUE, type = c("p", "a"), grid = TRUE)
```
